Modelling demography in the TREE package

### 2015-05-12 tree_demography.Rmd
Clean up vignettes (add refs, remove superfluous material). #7

# To do

- Table of units, symbols, subscripts
- Add equations for germination limit
- clean up text at start
- formula for EBT cohorts
- formula for error in EBT

@@

A system is physiologically or size structured whenever the basic rates of individuals (growth, mortality, reproduction) vary as functions of their size. These rates may also vary through environment, which can include feedbacks dependent on the current population state (e.g. density dependence).  Theory exists for tracing the development of a size- structured based on a partial differential equation (PDE) describing changes in the size distribution of individuals at any given time (describing the density function of individuals, rather than location of specific individuals). PSPs are formulated such that spatial distribution is neglected; only large populations are considered. Consequently the dynamics are entirely deterministic. These assumptions ensure similar individuals experience a homogeneous environment and are thus treated equally, and that inherently stochastic processes such as mortality can be reliably approximated by a deterministic description [makes use of the law of large numbers]. Individuals enter at the lower size class according to the rate of offspring production. The continuous nature of the system, with appropriate feedbacks, then ensures a dynamical solution to evolving size-structure including the problem of self-thinning. The dynamics of the population is given by the cumulative actions of its individuals.

## Patch- and size-structured population dynamics

@@

Such a system can be modelled as a metapopulation (sometimes called metacommunity for multiple species). Assuming patches are large, and there are many of them, the dynamics of the metapopulation can be captured deterministically using two PDEs (Kohyama 1993; deRoos 1997). Together these PDEs account for the combined influences of size-dependent variation in an individual's vital rates and disturbances on the size distribution of individuals. Below we outline equations suitable for modelling the equilibrium behaviour of such metapopulations.

The first PDE accounts for the effects of disturbances on the distribution of patch ages (defined as time since last disturbance) in the landscape (Von Foerster 1959; McKendrick 1926).

@@

This leads to a solution for the equilibrium distribution of patch-ages
\begin{equation} p(a) = p(0) \exp \left( \int_0^a \gamma(a^\prime) \; {\rm d} a^\prime \right),\end{equation}
where
\begin{equation}
\frac1{p(0)} = \left( \int_0^\infty \exp \left( \int_0^{a^{\prime\prime}} - \gamma(a^\prime) \; {\rm d} a^\prime \right) d a^{\prime\prime} \right),\end{equation}
 corresponds to the average lifespan of a patch.

The second PDE accounts for the development of competitive hierarchies within patches recovering from a disturbance (Sinko \& Streifer 1967; Hara 1984; Metz \& Diekmann 1986; de Roos 1997). Here we only consider cases where dispersal is restricted to newborns and where disturbances remove all established individuals within a patch. Let $n(x,m,a)$ represent the density of individuals with traits $x$ and mass $m$ in a patch of age $a$. For sufficiently large populations, the dynamics of $n$ are then given by the PDE

@@

### Environmental feedback

Individuals are assumed to interact  via their influence on the availability of resources. A competition kernel is required, giving the influence of an individual on the availability of resources for another individual. To get the total influence of all individuals on the target individual requires integration of the competition kernel over the density distribution $n$. If there are $N$ different species or types in the system, then their combined effects must be summed.

The approach outlined is very general and can be tailored to suit a range of different systems. To do so we need to describe:
\begin{itemize}
 \item the influence of an individual's size (state), its traits (type), and the availability of resources (environment) on the vital rates of growth, mortality and reproduction
 \item the dependence of resource availability on size-distribution of individuals within each species and their traits.
\end{itemize}

For plants, where we are interested in the effect of shading from other individuals on growth, mortality and fecundity, we can define a function for the amount of light at height $z$ in a patch of age $a$:

@@

Invasion fitness is most correctly defined as the long-term per capita growth rate of a rare mutant population growing in the environment determined by the resident strategy (Metz et al. 1992). Calculating per-capita growth rates, however, can be particularly challenging in a structured metapopulation model (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001). As an alternative measure of fitness, we can use the basic reproduction ratio, which gives the expected number of new dispersers arising from a single dispersal event. Evolutionary inferences made using the basic reproduction ratio will be similar to those made using per-capita growth rates for metapopulations at demographic equilibrium (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001).

Let $R\left(x^\prime,x\right)$ be the basic reproduction ratio of individuals with traits $x^\prime$ growing in the competitive environment of the resident traits $x$. Recalling that patches of age $a$ have density $p(a)$ in the landscape, it follows that any seed of $x^\prime$ has probability $p(a)$ of landing in a patch age $a$. The basic reproduction ratio for individuals with traits $x^\prime$ is then:
\begin{equation} \label{eq:InvFit}
R\left(x^\prime,x\right)=\int _0^{\infty }p\left(a\right)\tilde{R}\left(x^\prime,a, \infty \right)\; {\rm d}a ,
\end{equation}
where $\tilde{R}\left(x^\prime,a_0,a \right)$ is the expected number of dispersing offspring produced by a single disperser arriving in a patch of age $a_0$ up until age $a$ (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001). $\tilde{R}\left(x^\prime,a,\infty\right)$ is calculated by integrating an individual's fecundity over the expected lifetime the patch, taking into account competitive shading from residents with traits $x$, the individual's probability of surviving, and its traits via the equation:

\begin{equation} \label{eq:tildeR} \tilde{R}(x^\prime, a_0, a) =\int_{a_0}^{a}  \pi_0f(x^\prime, m(x^\prime, a_0, a^\prime), a^\prime) \, S_{\rm I} (x^\prime, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime.\end{equation}

@@

Summary statistics of the metapopulation are obtained by integrating over the density distribution, weighting by patch abundance $p(a)$. Firstly, the total number of individuals in the metapopulation is given by
\begin{equation}\hat{n}(x) = \int_{0}^{\infty} \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a \, {\rm d}m,\end{equation}
and the average density of plants size $m$ by
\begin{equation}\bar{n}(x,m) = \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a.\end{equation}

Average values for other quantities can also be calculated. Let $\phi(x, m, a)$ be a demographic quantity of interest, such growth rate, mortality rate, light environment, or size. The average value of $\phi$ across the metapopulation is
	\begin{equation}\hat{\phi}(x) = \frac1{\hat{n}(x) }\int_{0}^{\infty} \int_{0}^{\infty} p(a) \; n(x,m,a) \phi(x,m,a) \, {\rm d}a \, {\rm d}m,\end{equation}
while the average  value of $\phi$ for plants size $m$ is
 \begin{equation}\bar{\phi}(x,m) = \frac1{\bar{n}(x,m) }\int_{0}^{\infty}p(a)  n(x,m,a) \phi(x,m,a)\, {\rm d}a.\end{equation}

@@

The EBT solves the PDE describing development of $n(x,m,a)$ (Eq. $\ref{eq:PDE}$) by approximating the density function with a collection of cohorts spanning the size spectrum (de Roos 1988, 1997). Following a disturbance, a series of cohorts are introduced into each patch. These cohorts are then transported along the characteristics of Eq. $\ref{eq:PDE}$ according to the individual-level growth function. (Characteristics are curves along which Eq. $\ref{eq:PDE}$  becomes an ordinary differential equation (ODE); biologically, these are the growth trajectories of individuals, provided they do not die). The numerical technique consists of approximating changes to the number of individuals within each cohort by a system of closed-form linear ODEs. Eq. $\ref{eq:PDE}$ is thereby reduced to a family of ODEs, which can be stepped using an appropriate ODE solver with an adaptive step size (Press 1995).

Below we consider two alternative implementations of the EBT. Both require that we introduce a series of cohorts into a patch, and that these be transported along the characterstics of the Eq. $\ref{eq:PDE}$ (see figure).

<!-- ![Cohorts in the EBT](figs/EBT.png) -->

@@

## First method for approximating $n(x,m,a)$, based on modelling of cohort boundaries

@@

### Size

From equation  $\ref{eq:size}$, the boundary develops as

\begin{equation} \label{eq:boundSize} m_i(x, a_0, a) = m_0 + \int_{a_0}^{a} g(x,m_i(x, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime.\end{equation}

@@

### Survival
From equation  $\ref{eq:survivalIndividual}$, the survival of the individuals at the boundary individuals is

\begin{equation} \label{eq:boundSurv} S_{\rm I} (x, a_0, a) =  \pi_1 (x^\prime,m_0, a_0)  \exp\left(- \int_{a_0}^{a} d(x,m_i(x, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right).\end{equation}

@@

### Density of individuals

@@

\begin{equation} \label{eq:boundN} \begin{array}{ll}  n(x, m, a) & =n(x,m_0 ,a_0) \times \\
& \exp \left(-\int _{a_0}^{a} \left[\frac{\partial g(x,m(x, a_0, a^\prime),a^\prime)}{\partial m} +d(x,m(x, a_0, a^\prime),a^\prime)\right] {\rm d}a^\prime \right),\end{array} \end{equation}

@@

\begin{equation}\label{eq:boundN3}
n(x,m,a)=\frac{y_x}{ g(x,m) } \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x , m^\prime) }{g(x, m^\prime)} \, \textrm{d}m^\prime \bigg). \end{equation}
(for a proof, see \cite{de_roos_gentle_1997} page 153). Note also, that the integral on the RHS of $\ref{eq:boundN3}$ is simply the survival of individuals from $a_0 \rightarrow a$ describd in equation $\ref{eq:boundSurv}$
\begin{equation} \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x, m^\prime)}{g(x, m^\prime)}  \textrm{d} m^\prime \bigg) = \exp \bigg(-\int _{a_{0} }^{a}d(x, m(a^\prime)) \textrm{d}{a^\prime} \bigg) = S_{\rm I} (x, a_0, a).\end{equation}

@@

### Seed production

The lifetime seed production of boundary individuals  is calculated according to Eq. $\ref{eq:tildeR}$ as
\begin{equation}\begin{array}{ll}\tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\\end{array}\end{equation}
where $S_{\rm I}$ is individual survival (defined above) and
$S_{\rm P}$ is calculated as in Eq. $\ref{eq:survivalPatch}$.

@@

### Invasion fitness

@@

### Environmental feedback

@@

## Second method for approximating $n(x,m,a)$: based on modelling dynamics within cohorts

The original EBT, derived by de Roos (1997, 1988), proceeded by approximating the first and second moments of the density distribution $n \left( x,m,a \right)$ within sub-domains $\Omega_i$, represented by $\lambda_i$ and $\mu_i$, these being the total number and mean size of individuals within $\Omega_i$ respectively.

Under this assumption, the rates of change $\frac{\rm d} {{\rm d}a} \lambda_i$ and $\frac{\rm d} {{\rm d}a} \mu_i$ can be approximated given by two ODEs, which can be approximated by first-order closed-form solutions. These approximations form the centrepiece of the original EBT, and are derived as follows.

### First moment: Number of individuals

*Definition:*
\begin{equation} \label{eq:lambda}  \lambda_i = \int_{\Omega_i} n(x,m,a) \; {\rm d}m \end{equation}

\begin{equation} \label{eq:dlambda} \frac{{\rm d} }{{\rm d}a} \lambda_i = -\int_{\Omega_i} d(x,m,E) \; n(x,m,a) \; {\rm d}m \end{equation}

*Numerical technique:* The rate of change in $\lambda_i$ is approximated by

\begin{equation} \label{eq:dlambda_approx} \frac{{\rm d} }{{\rm d}a} \lambda_i =  d(x,\mu_i,E) \lambda_i \end{equation}, with $\lambda_i$ estimated by integrating this quantity with an ODE stepper. The approximation for \ref{eq:dlambda_approx} is only accurate to second order (for details on derivation see [appendix](#DeriveEBTapproximation)).

### Second moment: Average size

*Definition:*
\begin{equation}  \label{eq:mu} \mu_i = \frac1{\lambda_i}\int_{\Omega_i} m \; n(x,m,a) \; {\rm d}m  \end{equation}

\begin{equation} \label{eq:dmu} \frac{{\rm d} }{{\rm d}a} \mu_i = \frac1{\lambda_i} \int_{\Omega_i} \left[ g(x,m,E) - ( m-\mu_i ) d(x,m,a) \right] n(x,m,a) \; {\rm d}m  \end{equation}

*Numerical technique:* The rate of change in $\mu_i$ is approximated by
\begin{equation}  \label{eq:dmu_approx} \frac{{\rm d} }{{\rm d}a} \mu_i =  g(x,\mu_i,E), \end{equation}

with $\mu_i$ estimated by integrating this quantity with an ODE stepper. The approximation for \ref{eq:dmu_approx} is only accurate to second order (for details on derivation see [appendix](#DeriveEBTapproximation)).

### Density of individuals

*Definition:* $n(x,m,a)$ for $m \in \Omega_i$

*Numerical technique:* The EBT technique solves the PDE by approximating the changing size distribution $n(x,m,a)$ with a series of point masses with position and amplitude given by $\lambda_i$ and $\mu_i$. If we let $\delta(z)$ be a function for a point mass of size 1 localised at $z=0$, then

\begin{equation} \label{eq:n_delta}  n(x,m,a) \approx = \sum_{i=1}^{k} \lambda_i(a) \; \delta(m-\mu_i(a)). \end{equation}

The density function $n(x,m,a)$ within each sub-domain can also be estimated from its first moment using a discontinuous, piecewise-linear function:

\begin{equation} n_i \left( x,m,a \right) \approx \frac{\lambda_i}{m_i-m_{i-1}}, \end{equation}

although this requires that the position of the cohort boundaries are also known (see previous section for equations).

### Seed production
The lifetime seed production of individuals located at the cohort boundary is calculated according to Eq. $\ref{eq:tildeR}$ as
\begin{equation}\label{eq:boundR}\begin{array}{ll}\tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\\end{array}\end{equation}
where $S_{\rm I}$ is individual survival (defined above) and
$S_{\rm P}$ is calculated as in Eq. $\ref{eq:survivalPatch}$. Using the EBT approximation, the total number of sees produced within a cohort including individuals born in the interval $(a_1, a_2)$ and up to age $a$ is given by

\begin{equation} \int_{a_1}^{a_2} \tilde{R}(x, a_0, a) \; {\rm d}a_0 = f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) S_{\rm P} (a_0,a^\prime) \; {\rm d}a^\prime.  \end{equation}

Thus the total number of seeds produced by individuals within the cohort is given by

\begin{equation} \label{eq:cohR} \int_{a_1}^{\infty}  \int_{a_1}^{a_2} \tilde{R}(x, a_0, a) \; {\rm d}a_0 {\rm d}a = \int_{a_1}^{\infty} f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) \, S_{\rm P} (a_1,a^\prime).\end{equation}

*Numerical technique:* To solve equation $\ref{eq:cohR}$ we need solve the IVP:
$$\frac{dy}{dt} = f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) S_{\rm P} (a_1,a^\prime),$$
$$ y(a_1) = 0.$$

### Dynamics of the boundary cohort

A slightly modified set of equations is needed to track the moments of the smallest cohort, whose lower bound is fixed at $m_0$. Specifically, a new quantity $\Pi$, representing the total accumulated mass within this boundary cohort, is monitored in place of the second moment, which would otherwise be undefined whenever $\lambda _i=0$. Our approximations for the boundary cohort differ slightly to those of de Roos (1997, 1988) in that we Taylor expand around the centre of mass for the cohort instead of the lower boundary. Because relative growth rates often decline quickly with increasing size, use of lower boundary can overestimate the growth rate of the cohort.

#### Number of individuals

*Definition:*
\begin{equation} \lambda_1 = \int_{\Omega_1} n(x,m,a) \; {\rm d}m \end{equation}

\begin{equation} \frac{{\rm d} }{{\rm d}a} \lambda_1 = y_x \pi_0 \pi_1(x,m_0, a) - \int_{\Omega_1} d(x,m,a) \; n(x,m,a) \; {\rm d}m  \end{equation}
(TODO: I $\pi_0$ in right place here? Should appear in seed production, not inflow).

*Numerical technique:* $\frac{{\rm d} }{{\rm d}a} \lambda_1 = y_x - d(x,\mu_1,a) \lambda_1$

#### New mass

*Definition:* $\Pi_1 = \int_{\Omega_1} (m-m_0) \; n(x,m,a) \; {\rm d}m$

*Exact:* $\frac{{\rm d} }{{\rm d}a} \Pi_1 = \int_{\Omega_1} \left[ g(x,m,E) - ( m-\mu0) d(x,m,a) \right] n(x,m,a) \; {\rm d}m$

*Numerical technique:* $\frac{{\rm d} }{{\rm d}a} \Pi_1 =
 g(x,\mu_1,a) \lambda_1 - \Pi_1  d(x,\mu_1,a)$

#### Average size

*Definition:*
\begin{equation}  \mu_1 = \left\{ \begin{array}{ll} \frac1{\lambda_i}\int_{\Omega_i} m \; n(x,m,a) \; {\rm d}m   & \textrm{if } \lambda_1 > 0 \\
m_0 & \textrm{otherwise}  \end{array} \right. \end{equation}

\begin{equation} \mu_1 = \left\{ \begin{array}{ll}  m_0 + \frac{\Pi_1}{\lambda_1}  & \textrm{if } \lambda_1 > 0 \\
m_0 & \textrm{otherwise}  \end{array} \right. \end{equation}

*Numerical technique:* NA, use exact.

### Environmental feedback

The estimated density function (Eq. \ref{eq:n_delta}) is used to calculate the amount of shading on each individual from other individuals in the patch, as described in Eq. \ref{eq:light}.

*Numerical technique:*  The approximation of  $n(x,m,a)$ by a series of delta peaks  in Eq. \ref {eq:n_delta} allows integration in Eq. \ref{eq:light} to be approximated as a sum over cohorts:

\begin{equation} \label{eq:light_deltasum}
E(z,a) = \exp \left(-c_{ext}  \sum_{n=1}^{N} \sum_{\Omega_i} \phi(\mu_{n,i}) \, L(z, h(\mu_{n,i})) \lambda_{n,i} \right), \end{equation}

where $\mu_{n,i}, \lambda_{n,i}$ are the average size and number of individuals within the i$^{th}$ cohort of species $n$ (Eqs. \ref{eq:mu}, \label{eq:lambda} respectively).

## Controlling error in the EBT

Errors in the EBT approximation to $n$ can arise from two sources: (i) poor spacing of cohorts in the size dimension, and (ii) when stepping cohorts through time. Error of the latter type is effectively controlled using a suitable ODE solver and is not considered further here. However, there is no existing method to control error arising from poor spacing of cohorts. Poor cohort spacing introduces error both because the equations governing cohort dynamics are only accurate up to second order (Table 4-1), and because any integrals over the size distribution (e.g. total offspring production, competitive effect) may be poorly resolved.

@@

Consider patches of habitat which are subject to some intermittent disturbance and where the age of a patch is corresponds to the time since the last disturbance. Let $p(a,t)$ be the frequency-density of patches age $a$ at time $t$ and let $\gamma(a)$ be the age-dependent probability that a patch of age $a$ is transformed into a patch of age $0$ through disturbance. Then according to the Von Foerster equation for age-structured population dynamics (von_foerster_remarks_1959), the dynamics of $p(a,t)$ are given by
$$ \frac{\partial }{\partial t} p(a,t)=-\frac{\partial }{\partial a} p(a,t)-\gamma(a,t)p(a,t),$$
with boundary condition
$$ p(0,t)=\int^{\infty}_{0}\gamma(a,t)p(a,t)\,{\rm d}a.$$

@@

For general notes on structured population and analysis of data see (Metz and Diekmann 1986, Taljapurkar and Caswell 1997, Diekmann et al. 1998, Diekmann et al. 2001)

@@

\begin{equation}\frac{\partial }{\partial t} \int _{m_{0} }^{m_{\infty } }n(m,a) \partial m=g(m_{0} ,a) \; n(m_{0} ,a)-g(m_{\infty } ,a) \; n(m_{\infty } ,a)-\int _{m_{0} }^{m_{\infty } }d (m,a) \; n(m,a) \partial m.\end{equation}

The LHS of this relationship is evidently the rate of change of total numbers of individual in the population, while the right-hand-term is the total population death rate. Further, $n(m_{\infty } ,a)=0$. Thus to balance total births and deaths, $g(m_{0} ,a) \; n(m_{0} ,a)$ must equal the birth rate $B(x, n, E,a)$. Thus the boundary condition is given by
\begin{equation}g(m_{0} ,a) \; n(m_{0} ,a)=B(x, n, E,a).\end{equation}

@@

## Exact solutions: Integration along characteristics {#IntegrationAlongCharacteristics}

Characteristics are the projections of an individual's trajectory in i-space and time, provided they do not die. By integrating along characteristics, the PDE can be solved to yield the a general solution for the density of individuals with time of birth $t_{b} > 0$:
\begin{equation}\label{eq:n1}
n(m,t)=n(m_0 , t_b) \exp \bigg(-\int _{t_b}^{t} \left[\frac{\partial g(m(\theta ),E(\theta ))}{\partial m} +d(m(\theta ),E(\theta ))\right] d\theta \bigg)
\end{equation}

This states that the density $n$ at a specific time is a product of density at the origin adjusted for changes through development and mortality. Using the second boundary condition above, $n(m_{0} ,t_{g})$ can be replaced by $B(x, n, E,t_{b}) / g(m_{0} ,E(t_{b}))$. Note that density decreases through time because of mortality (as per typical survival equation), but that growth also contributes to changes in density. If growth is slowing with size, (i.e. $\frac{\partial g(m(\theta ),E(\theta ))}{\partial m}  <0$), then density increases because the characteristics converge on one another. Eq. \ref{eq:n1} can be solved (numerically or analytically) to give density with full ecological feedback and in changing environemnts. \\

In cases where the environment is constant and time-independent ($E(t)=E_c$), it is possible to further simplify Eq. \ref{eq:n1} to give (see \cite{de_roos_gentle_1997} page 153 for proof):

>>>>

@@

@@

Such a system can be modelled as a metapopulation (sometimes called metacommunity for multiple species). Assuming patches are large, and there are many of them, the dynamics of the metapopulation can be captured deterministically using two PDEs [@Kohyama-1993; @Deroos-1997; @Moorcroft-2001]. Together these PDEs account for the combined influences of size-dependent variation in an individual's vital rates and disturbances on the size distribution of individuals. Below we outline equations suitable for modelling the equilibrium behaviour of such metapopulations.

The first PDE accounts for the effects of disturbances on the distribution of patch ages (defined as time since last disturbance) in the landscape [@Vonfoerster-1959; @Mckendrick-1926].

@@

This leads to a solution for the equilibrium distribution of patch-ages
\begin{equation} p(a) = p(0) \exp \left( \int_0^a \gamma(a^\prime) \; {\rm d} a^\prime \right),
\end{equation}
where
\begin{equation}
  \frac{p(0)} = \left( \int_0^\infty \exp \left( \int_0^{a^{\prime\prime}} - \gamma(a^\prime) \; {\rm d} a^\prime \right) d a^{\prime\prime} \right),
\end{equation}
 corresponds to the average lifespan of a patch.

The second PDE accounts for the development of competitive hierarchies within patches recovering from a disturbance [@Sinko-1967;  @Hara-1984; @Metz-1986; @Deroos-1997]. Here we only consider cases where dispersal is restricted to newborns and where disturbances remove all established individuals within a patch. Let $n(x,m,a)$ represent the density of individuals with traits $x$ and mass $m$ in a patch of age $a$. For sufficiently large populations, the dynamics of $n$ are then given by the PDE

@@

## Environmental feedback

Individuals are assumed to interact  via their influence on the availability of resources. For plants, where we are interested in the effect of shading from other individuals on growth, mortality and fecundity, we can define a function for the amount of light at height $z$ in a patch of age $a$:

@@

Invasion fitness is most correctly defined as the long-term per capita growth rate of a rare mutant population growing in the environment determined by the resident strategy [@Metz-1992]. Calculating per-capita growth rates, however, can be particularly challenging in a structured metapopulation model [@Gyllenberg-2001; @Metz-2001]. As an alternative measure of fitness, we can use the basic reproduction ratio, which gives the expected number of new dispersers arising from a single dispersal event. Evolutionary inferences made using the basic reproduction ratio will be similar to those made using per-capita growth rates for metapopulations at demographic equilibrium [@Gyllenberg-2001; @Metz-2001].

Let $R\left(x^\prime,x\right)$ be the basic reproduction ratio of individuals with traits $x^\prime$ growing in the competitive environment of the resident traits $x$. Recalling that patches of age $a$ have density $p(a)$ in the landscape, it follows that any seed of $x^\prime$ has probability $p(a)$ of landing in a patch age $a$. The basic reproduction ratio for individuals with traits $x^\prime$ is then:
\begin{equation} \label{eq:InvFit}
  R\left(x^\prime,x\right)=\int _0^{\infty }p\left(a\right)\tilde{R}\left(x^\prime,a, \infty \right)\; {\rm d}a ,
\end{equation}
where $\tilde{R}\left(x^\prime,a_0,a \right)$ is the expected number of dispersing offspring produced by a single disperser arriving in a patch of age $a_0$ up until age $a$ [@Gyllenberg-2001; @Metz-2001]. $\tilde{R}\left(x^\prime,a,\infty\right)$ is calculated by integrating an individual's fecundity over the expected lifetime the patch, taking into account competitive shading from residents with traits $x$, the individual's probability of surviving, and its traits via the equation:

\begin{equation} \label{eq:tildeR}
  \tilde{R}(x^\prime, a_0, a) =\int_{a_0}^{a}  \pi_0f(x^\prime, m(x^\prime, a_0, a^\prime), a^\prime) \, S_{\rm I} (x^\prime, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime.
\end{equation}

@@

Summary statistics of the metapopulation are obtained by integrating over the density distribution, weighting by patch abundance $p(a)$. Firstly, the total number of individuals in the metapopulation is given by
\begin{equation}
  \hat{n}(x) = \int_{0}^{\infty} \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a \, {\rm d}m,
\end{equation}
and the average density of plants size $m$ by
\begin{equation}\bar{n}(x,m) = \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a.
\end{equation}

Average values for other quantities can also be calculated. Let $\phi(x, m, a)$ be a demographic quantity of interest, such growth rate, mortality rate, light environment, or size. The average value of $\phi$ across the metapopulation is
\begin{equation}\hat{\phi}(x) = \frac1{\hat{n}(x) }\int_{0}^{\infty} \int_{0}^{\infty} p(a) \; n(x,m,a) \phi(x,m,a) \, {\rm d}a \, {\rm d}m,
\end{equation}
while the average  value of $\phi$ for plants size $m$ is
\begin{equation}\bar{\phi}(x,m) = \frac1{\bar{n}(x,m) }\int_{0}^{\infty}p(a)  n(x,m,a) \phi(x,m,a)\, {\rm d}a.
\end{equation}

@@

Our approach for modelling solving size-structured population dynamics is based on the Escalator Boxcar Train technique  (EBT) [@Deroos-1997; @Deroos-1992; @Deroos-1988]. The EBT solves the PDE describing development of $n(x,m,a)$ (Eq. $\ref{eq:PDE}$) by approximating the density function with a collection of cohorts spanning the size spectrum. Following a disturbance, a series of cohorts are introduced into each patch. These cohorts are then transported along the characteristics of Eq. $\ref{eq:PDE}$ according to the individual-level growth function. (Characteristics are curves along which Eq. $\ref{eq:PDE}$  becomes an ordinary differential equation (ODE); biologically, these are the growth trajectories of individuals, provided they do not die).

The original EBT [@Deroos-1997; @Deroos-1992; @Deroos-1988] proceeds by approximating the first and second moments of the density distribution $n \left( x,m,a \right)$ within sub-domains $\Omega_i$, represented by $\lambda_i$ and $\mu_i$, these being the total number and mean size of individuals within $\Omega_i$ respectively. Under this assumption, the rates of change $\frac{\rm d} {{\rm d}a} \lambda_i$ and $\frac{\rm d} {{\rm d}a} \mu_i$ can be approximated given by two ODEs, which can in turn be approximated by first-order closed-form solutions. Eq. $\ref{eq:PDE}$ is thereby reduced to a family of ODEs, which can be stepped using an appropriate ODE solver with an adaptive step size (Press 1995). The size distribution $n(x,m,a)$ is then approximated by a series of point masses with position and amplitude given by $\lambda_i$ and $\mu_i$.

In the current implementation we use a different approach for approximating the size distribution $n(x,m,a)$. Instead of tracking the first and second moments of the size-distribution within cohorts, we track a point mass estimate of $n$ along the characteristics corresponding to cohort boundary. By integrating along characteristics, the PDE can be solved to yield the a general solution for the density of individuals with time of birth $t_{b} > 0$:
\begin{equation}\label{eq:n1}
  n(m,t)=n(m_0 , t_b) \exp \bigg(-\int _{t_b}^{t} \left[\frac{\partial g(m(\theta ),E(\theta ))}{\partial m} +d(m(\theta ),E(\theta ))\right] d\theta \bigg)
\end{equation}
Eq. \ref{eq:n1} states that the density $n$ at a specific time is a product of density at the origin adjusted for changes through development and mortality. Using the second boundary condition above, $n(m_{0} ,t_{g})$ can be replaced by $B(x, n, E,t_{b}) / g(m_{0} ,E(t_{b}))$. Note that density decreases through time because of mortality (as per typical survival equation), but that growth also contributes to changes in density. If growth is slowing with size, (i.e. $\frac{\partial g(m(\theta ),E(\theta ))}{\partial m}  <0$), then density increases because the characteristics converge on one another. In cases where the environment is constant and time-independent, it may be possible to further simplify Eq. \ref{eq:n1} to give an analytical solution  for $n$ (see [@Deroos-1997} page 153 for proof).  In most cases involving interesting ecological feedback, however, eq. \ref{eq:n1} must be solved numerically.

@@

@@

## Size

From equation  $\ref{eq:size}$, the boundary develops as
\begin{equation} \label{eq:boundSize}
  m_i(x, a_0, a) = m_0 + \int_{a_0}^{a} g(x,m_i(x, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime.
\end{equation}

@@

## Survival
From equation  $\ref{eq:survivalIndividual}$, the survival of the individuals at the boundary individuals is

\begin{equation} \label{eq:boundSurv} S_{\rm I} (x, a_0, a) =  \pi_1 (x^\prime,m_0, a_0)  \exp\left(- \int_{a_0}^{a} d(x,m_i(x, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right).
\end{equation}

@@

## Density of individuals

@@

\begin{equation} \label{eq:boundN}
  \begin{array}{ll}
  n(x, m, a) & =n(x,m_0 ,a_0) \times \\
  & \exp \left(-\int _{a_0}^{a} \left[\frac{\partial g(x,m(x, a_0, a^\prime),a^\prime)}{\partial m} +d(x,m(x, a_0, a^\prime),a^\prime)\right] {\rm d}a^\prime \right),
  \end{array}
\end{equation}

@@

\begin{equation}\label{eq:boundN3}
	n(x,m,a)=\frac{y_x}{ g(x,m) } \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x , m^\prime) }{g(x, m^\prime)} \, \textrm{d}m^\prime \bigg).
\end{equation}
(for a proof, see [@Deroos-1997] page 153). Note also, that the integral on the RHS of $\ref{eq:boundN3}$ is simply the survival of individuals from $a_0 \rightarrow a$ describd in equation $\ref{eq:boundSurv}$
\begin{equation} \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x, m^\prime)}{g(x, m^\prime)}  \textrm{d} m^\prime \bigg) = \exp \bigg(-\int _{a_{0} }^{a}d(x, m(a^\prime)) \textrm{d}{a^\prime} \bigg) = S_{\rm I} (x, a_0, a).
\end{equation}

@@

## Seed production

The lifetime seed production of boundary individuals  is calculated according to Eq. $\ref{eq:tildeR}$ as
\begin{equation}
  \begin{array}{ll}
  \tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\
  \end{array}
\end{equation}
where $S_{\rm I}$ is individual survival (defined above) and
$S_{\rm P}$ is calculated as in Eq. $\ref{eq:survivalPatch}$.

@@

@@

## Environmental feedback

@@

# Controlling error in the EBT

Errors in the EBT approximation to $n$ can arise from two sources: (i) poor spacing of cohorts in the size dimension, and (ii) when stepping cohorts through time. Error of the latter type is effectively controlled using a suitable ODE solver and is not considered further here. However, there is no existing method to control error arising from poor spacing of cohorts. Poor cohort spacing introduces error both because the equations governing cohort dynamics are only accurate up to second order, and because any integrals over the size distribution (e.g. total offspring production, competitive effect) may be poorly resolved.

@@

Consider patches of habitat which are subject to some intermittent disturbance and where the age of a patch is corresponds to the time since the last disturbance. Let $p(a,t)$ be the frequency-density of patches age $a$ at time $t$ and let $\gamma(a)$ be the age-dependent probability that a patch of age $a$ is transformed into a patch of age $0$ through disturbance. Then according to the Von Foerster equation for age-structured population dynamics [@Vonfoerster-1959], the dynamics of $p(a,t)$ are given by
$$ \frac{\partial }{\partial t} p(a,t)=-\frac{\partial }{\partial a} p(a,t)-\gamma(a,t)p(a,t),$$
with boundary condition
$$ p(0,t)=\int^{\infty}_{0}\gamma(a,t)p(a,t)\,{\rm d}a.$$

@@

@@

\begin{equation}\frac{\partial }{\partial t} \int _{m_{0} }^{m_{\infty } }n(m,a) \partial m=g(m_{0} ,a) \; n(m_{0} ,a)-g(m_{\infty } ,a) \; n(m_{\infty } ,a)-\int _{m_{0} }^{m_{\infty } }d (m,a) \; n(m,a) \partial m.
\end{equation}

The LHS of this relationship is evidently the rate of change of total numbers of individual in the population, while the right-hand-term is the total population death rate. Further, $n(m_{\infty } ,a)=0$. Thus to balance total births and deaths, $g(m_{0} ,a) \; n(m_{0} ,a)$ must equal the birth rate $B(x, n, E,a)$. Thus the boundary condition is given by
\begin{equation}g(m_{0} ,a) \; n(m_{0} ,a)=B(x, n, E,a).
\end{equation}

@@
